---
title: "Regression models - Course project - Motor Trend Analysis"
date: "01/16/2015"
output: pdf_document
---

In this report we try to addres the  the following two questions:

* “Is an automatic or manual transmission better for MPG”
* "Quantify the MPG difference between automatic and manual transmissions"

We have found that manual transmission overall, seems to give a better mpg than
automatic transmission. In this report we show though, that other variables such
as number of cylinders, weight and horsepower, play an important role as well.

# Sanity check

First we look at the data to have an idea what we're looking at. We see that,
next to mpg, we have `r ncol(mtcars) - 1` numeric columns. 

```{r, results="hide"}
data(mtcars)
str(mtcars)  # See appendix for results
```

However, some of them should be treated as factor:

```{r}
mtcars$cyl <- factor(mtcars$cyl)
mtcars$vs <- factor(mtcars$vs, labels=c("V-engine", "Straight engine"))
mtcars$gear <- factor(mtcars$gear)
mtcars$carb <- factor(mtcars$carb)
mtcars$am <- factor(mtcars$am,labels=c('Automatic','Manual'))
```

# Exploratory analysis

We draw a pairs plot to have a first look at the relation ship between mpg and
the other variables (see appendix). There we already see some trends, such as

* An increase in number of cylinders tends to lead to a decrease in mpg.
* An increase in hp tends to lead to a decrease in mpg.
* An increase in rear axle ratio tends (though more varition) to lead to an increase in mpg.
* An increase in weight, seems to lead to a decrease in mpg

# Model fitting

We now fit a linear model, initially using all variables and see if we get
results which are usable. When looking at the summary, we see that we get high
p-values for all variables, suggesting that this model is not such a good fit.
We choose a better model by using a step wise algorithm, in backward direction
(i.e. remove candidates from the fit).

```{r, results="hide"}
vars.all <- lm(mpg ~ ., data = mtcars)
vars.best.fit <- step(vars.all, direction = "backward")
```

```{r}
summary(vars.best.fit)
```

**Interpretation:** From this model's summary we see that the structure in the
data is reasonably well explained by the variables (or variable values in case
of categorical variables): cylinder 16, cylinder 18, horsepower, weight and manual
transmision. Most of those variables have pretty low p-values (< 0.05), which increases
our confidence in the model. Additionally, we see that the R-squared values are
pretty high, that is around 84 percent of the in- or decrease in milage per gallon
is explained by our model. Finally, we see that the 1st and 3rd quantiles of the
residuals are about the same as `1.5 +/- std error` given a strong hint that the
residuals are normally distributed. This observation is further strengthened by
the normal q-q plot (see appendix), though there some outliers become apparent.

From this model, we see that we can expect an `1.8092` increase in mpg when
changing from automatic to manual, keeping all other variables fixed. This
suggests that having a manual car is better for mpg. To further test this
hypothesis, we perform a t-test:

```{r, results="hide"}
t.test(mpg ~ am, data = mtcars) # See appendix
```

Finally, we construct a 95% confidence interval to find the values below. This
still includes a possible decrease, which shows that just switching from
automatic to manual is not to be sure (with earlier mentioned confidence) that
it will actually increase milage.

```{r}
confint(vars.best.fit)["amManual",]
```

# Appendix

```{r}
str(mtcars)
```

```{r, results="hide"}
pairs(mpg ~ ., data=mtcars)
```

```{r, results="hide"}
par(mfrow=c(2,2))
plot(vars.best.fit)
```

```{r}
t.test(mpg ~ am, data = mtcars) # See appendix
```
