## Reproducible Research: Peer Assessment 1

### Loading and preprocessing the data

We first unzip the data and read it into R.

```{r}
unzip("activity.zip")
activity <- read.csv("activity.csv")
```

Next we process the data to make it suitable for further analysis. In this case
it means properly parsing the date string as a Date object.

```{r}
activity$date <- as.Date(activity$date, format="%Y-%m-%d")
```

### What is mean total number of steps taken per day?

For this part of the analysis, we ignore the missing values in the dataset.

```{r}
activityNoNa <- activity[!is.na(activity$steps),]
```

Next, we summarize the data to get steps per day. 

```{r}
stepsPerDay <- rowsum(activityNoNa, activityNoNa$date)
stepsPerDay$date <- as.Date(row.names(stepsPerDay), format="%Y-%m-%d")
```
Let's see how the steps per day have evolved over time by plotting a histogram.

```{r}
library(ggplot2)
stepsPerDayPlot <- ggplot(stepsPerDay, aes(x = date, y = steps))
stepsPerDayPlot <- stepsPerDayPlot + geom_bar(stat = "identity")
stepsPerDayPlot
```

As we can see in the plot, the number of steps is quite differring. There are
various days on which no steps where recorded at all. A peak can be found around
the end of November.

When we have a look at the **mean** and **median** total number of steps taken
per day:

```{r}
mean(stepsPerDay$steps)
median(stepsPerDay$steps)
```

we see that they don't differ a lot. Suggesting that the distribution is quite
symetric. This is confirmed when we look at the histogram:

```{r}
qplot(stepsPerDay$step, geom="bar")
```

### What is the average daily activity pattern?

To answer this question we first make a time series plot of the 5-minute
interval (x-axis) and the average number of steps taken, averaged across all
days (y-axis):

```{r}
stepsPerInterval <- aggregate(activityNoNa, list(interval = activityNoNa$interval), mean)
stepsPerInterval$time <- formatC(stepsPerInterval$interval, width = 4, format = "d", flag = "0")
stepsPerInterval$time <- as.POSIXct(stepsPerInterval$time, format="%H%M")
stepsPerIntervalPlot <- ggplot(stepsPerInterval, aes(x = time, y = steps)) +
   geom_bar(stat = "identity")
stepsPerIntervalPlot
```

We can get the 5-minute interval, that on average across all the days in the
dataset, contains the maximum number of steps as follows:

```{r}
mostStepsOfTheDay <- stepsPerInterval[which.max(stepsPerInterval$steps),]$interval
```

That is, this person, on average takes the most steps at `r mostStepsOfTheDay`.

### Imputing missing values

Note that there are a number of days/intervals where there are missing values
(coded as `NA`). The presence of missing days may introduce bias into some
calculations or summaries of the data.

Therefore we calculate and report the total number of missing values in the
dataset (i.e. the total number of rows with `NA`s):

```{r}
sum(is.na(activity$steps))
```

Next, we devise a strategy for filling in all of the missing values in the
dataset:

As we have already calculated the mean value for each interval, we will reuse
these values to fill in the gaps.

Now we have chosen a strategy, we create a new dataset that is equal to the
original dataset but with the missing data filled in.

```{r}
activityImputed <- NULL
activityMissing <- activity[is.na(activity$steps), ]

for(i in 1:nrow(activityMissing)) {
    row <- activityMissing[i,]
    row$steps <- stepsPerInterval[stepsPerInterval$interval == row$interval, ]$steps[1]
    activityImputed <<- rbind(row, activityImputed)
}

activityImputed <- rbind(activityImputed, activity[!is.na(activity$steps), ])
```

We reuse our code from before to ake a histogram of the total number of steps
taken each day.

```{r}
stepsPerDayImp <- rowsum(activityImputed, activityImputed$date)
stepsPerDayImp$date <- as.Date(row.names(stepsPerDayImp), format="%Y-%m-%d")

stepsPerDayImpPlot <- ggplot(stepsPerDayImp, aes(x = date, y = steps))
stepsPerDayImpPlot <- stepsPerDayImpPlot + geom_bar(stat = "identity")
stepsPerDayImpPlot
```

Next, we calculate and report the **mean** and **median** total number of steps
taken per day.

```{r}
mean(stepsPerDayImp$steps)
median(stepsPerDayImp$steps)
```

As we can see, this hardly changes the values which we calculated before. The
most 'striking' difference is that now the median and the average steps per day
are exactly the same.

### Are there differences in activity patterns between weekdays and weekends?
