## Reproducible Research: Peer Assessment 1

### Loading and preprocessing the data

We first unzip the data and read it into R.

```{r}
unzip("activity.zip")
activity <- read.csv("activity.csv")
```

Next we process the data to make it suitable for further analysis. In this case
it means properly parsing the date string as a Date object.

```{r}
activity$date <- as.Date(activity$date, format="%Y-%m-%d")
```

### What is mean total number of steps taken per day?

For this part of the analysis, we ignore the missing values in the dataset.

```{r}
activityNoNa <- activity[!is.na(activity$steps),]
```

Next, we summarize the data to get steps per day. 

```{r}
stepsPerDay <- rowsum(activityNoNa, activityNoNa$date)[, c(1,2)]
stepsPerDay$date <- as.Date(row.names(stepsPerDay), format="%Y-%m-%d")
```
Let's see how the steps per day have evolved over time by plotting a histogram.

```{r}
library(ggplot2)
stepsPerDayPlot <- ggplot(stepsPerDay, aes(x = date, y = steps))
stepsPerDayPlot <- stepsPerDayPlot + geom_bar(stat = "identity")
stepsPerDayPlot
```

As we can see in the plot, the number of steps is quite differring. There are
various days on which no steps where recorded at all. A peak can be found around
the end of November.

When we have a look at the **mean** and **median** total number of steps taken
per day:

```{r}
mean(stepsPerDay$steps)
median(stepsPerDay$steps)
```

we see that they don't differ a lot. Suggesting that the distribution is quite
symetric. This is confirmed when we look at the histogram:

```{r}
qplot(stepsPerDay$step, geom="bar")
```

### What is the average daily activity pattern?



### Imputing missing values



### Are there differences in activity patterns between weekdays and weekends?
